---
title: "Linear Regression: Basics"
description: "My take on machine learning based on statistics"
author: "Karl Marquez"
date: 2025-08-20
categories:
  - machine learning
  - r
  - statistics
format: 
  html: 
    code-fold: true
    code-summary: "Show me the code"
    code-overflow: wrap
    css: styles.css
execute: 
  warning: false
---

To tackle linear regression, let's set up the environment and upload the data advertising data set: 

```{r}
packages <- c("ISLR2", "tidyverse", "ggplot2", "gridExtra", "plotly",
              "scatterplot3d")
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
invisible(lapply(packages, library, character.only = TRUE))

library(tidyverse)
library(ISLR2)
library(ggplot2)
library(gridExtra)
library(plotly)
library(scatterplot3d)

karl_theme <- theme_bw() +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        legend.title = element_text(size=12),
        legend.text = element_text(size=12))

advertising <- read.csv("Advertising.csv")
```

Using Advertising data set, shown is a scatterplot of three advertising methods (input) to sales (output).

```{r}
#| fig-cap: "TV, radio, and newspaper seems to have a positive association to sales. The more the company spends on TV advertisement, the higher the associated sales. Newspaper seems to be the weakest advertising method."

p1 <- ggplot(data = advertising, aes(x = TV, y = sales)) +
        geom_point(alpha = 0.75, size = 2, color = "steelblue") +
        geom_smooth(method = "lm") +
        karl_theme
p2 <- ggplot(data = advertising, aes(x = radio, y = sales)) +
        geom_point(alpha = 0.75, size = 2, color = "steelblue") +
        geom_smooth(method = "lm") +
        karl_theme
p3 <- ggplot(data = advertising, aes(x = newspaper, y = sales)) +
        geom_point(alpha = 0.75, size = 2, color = "steelblue") +
        geom_smooth(method = "lm") +
        karl_theme
grid.arrange(p1, p2, p3, ncol = 3)

```

Some questions that linear regression can answer: 
1. Is there a relationship between advertising budget and sales? 
2. How strong is the relationship between advertising budget and sales?  
3. Which media are associated with sales? 
4. How large is the association between each medium and sales? 
5. How accurately can we predict future sales? 











