---
title: "Linear Regression: Basics"
description: "My take on machine learning based on statistics"
author: "Karl Marquez"
date: 2025-08-20
categories:
  - machine learning
  - r
  - statistics
format: 
  html: 
    code-fold: true
    code-summary: "Show me the code"
    code-overflow: wrap
    css: styles.css
execute: 
  warning: false
---

To tackle linear regression, let's set up the environment and upload the data advertising data set: 

```{r}
packages <- c("ISLR2", "tidyverse", "ggplot2", "gridExtra", "plotly",
              "scatterplot3d")
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
invisible(lapply(packages, library, character.only = TRUE))

library(tidyverse)
library(ISLR2)
library(ggplot2)
library(gridExtra)
library(plotly)
library(scatterplot3d)

karl_theme <- theme_bw() +
  theme(plot.title = element_text(size=20),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12),
        legend.title = element_text(size=12),
        legend.text = element_text(size=12))

advertising <- read.csv("Advertising.csv")
```

Using Advertising data set, shown is a scatterplot of three advertising methods (input) to sales (output).

```{r}
#| fig-cap: "TV, radio, and newspaper seems to have a positive association to sales. The more the company spends on TV advertisement, the higher the associated sales. Newspaper seems to be the weakest advertising method."

p1 <- ggplot(data = advertising, aes(x = TV, y = sales)) +
        geom_point(alpha = 0.75, size = 2, color = "steelblue") +
        geom_smooth(method = "lm") +
        karl_theme
p2 <- ggplot(data = advertising, aes(x = radio, y = sales)) +
        geom_point(alpha = 0.75, size = 2, color = "steelblue") +
        geom_smooth(method = "lm") +
        karl_theme
p3 <- ggplot(data = advertising, aes(x = newspaper, y = sales)) +
        geom_point(alpha = 0.75, size = 2, color = "steelblue") +
        geom_smooth(method = "lm") +
        karl_theme
grid.arrange(p1, p2, p3, ncol = 3)

```

Some questions that linear regression can answer: 

1. Is there a relationship between advertising budget and sales? 
2. How strong is the relationship between advertising budget and sales?  
3. Which media are associated with sales? 
4. How large is the association between each medium and sales? 
5. How accurately can we predict future sales? 
6. Is the relationship linear? 
7. Is there synergy among the advertising media? 

## Simple Linear Regression 
It is a method of predicting a quantitative $Y$ on the basis of a single predictor variable $X$, assuming the relationship between $X$ and $Y$ is linear. It is given by the equation 
$$
Y \approx \beta_0 + \beta_1 X
$$
In our advertising data set, $Y$ is sales, and $X$ can be TV. We can regress sales onto TV by fitting the model 
$$
\text{sales} \approx \beta_0 + \beta_1 \times \text{TV}
$$
Here, the unknown constants, or coefficient parameters, are $\beta_0$ (*intercept*) and $\beta_1$ (*slope*). We estimate these numbers using the training data set. Once we have the coefficients, we can predict future sales using the same equation 
$$
\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x
$$
The *hat* symbol, ^, is the estimated value for an unknown parameter, coefficient, or predicted response. 

### Estimating the Coefficients 
The question is how do we calculate the coefficient parameters using the training data set? We want to find the $\beta_0$ (*intercept*) and $\beta_1$ (*slope*) such that the resulting line is as close as possible to the data points. How do we measure this *closeness*? The most common approach is the *least squares* criterion. To illustrate, see below:
```{r}
advertising <- advertising %>% 
  mutate(res = residuals(loess(sales ~ TV)))
ggplot(data = advertising, aes(x = TV, y = sales)) +
  geom_point(alpha = 0.75, size = 2.5, color = "steelblue") +
  geom_smooth(method = lm, color = "blue") +
  geom_segment(aes(xend = TV, yend = sales - res), color = "red", alpha = 0.75) +
  karl_theme

```








